MRD_microbiome_working

### 2019-02-06 MRD_microbiome_working.Rmd
updated

```{r cars}

library(microbiome)
library(knitr)
library(ggpubr)
library(dplyr)

# Data from http://dx.doi.org/10.1038/ncomms7342
data(atlas1006)   
# Rename the example data (which is a phyloseq object)
pseq <- atlas1006

data(dietswap)  
pseqds <- dietswap
print(dietswap)

@@

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}

# Richness
tab2 <- richness(pseqds)
kable(head(tab2))

@@

```{r , echo=FALSE}

# Richness
tab2 <- richness(pseqds)
kable(head(tab2))

@@

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

>>>>

```{r data set-up}

setwd("/Users/rezahaji-dehabadi/scripts/workspace")

	## Get data
	library(foreign)
	library(stargazer)
	library(psych)
	library(plyr)
	library(xtable)
	library(car)
	#library(dplyr)
	library(scales)
	library(reshape) 
	library(gmodels)
	library(nnet)
	library(tidyr)
	library(sjmisc)
	library(RColorBrewer)
  library(htmltools)
	library(broom)

library(hrbrthemes)
library(gcookbook)
library(tidyverse)
library(gdtools)
library(phyloseq)
library(microbiome)
library(knitr)
library(ggpubr)

library(phylogeo)
library(dplyr)

# Data from http://dx.doi.org/10.1038/ncomms7342
data(atlas1006)

# Rename the example data (which is a phyloseq object)
pseq <- atlas1006

@@

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

```{r summary}

# Summarise contents of phyloseq
summarize_phyloseq(pseq)

```{r alpha diversity}

# Alpha diversity
ps1 <- prune_taxa(taxa_sums(pseq) > 0, pseq)
tab <- diversities(ps1, index = "all")
stargazer(tab, type = "text",summary = NULL)
kable(head(tab))
plot(tab)

```{r richness}

# Richness
tab <- richness(pseq)
kable(head(tab))
stargazer(tab, type = "text",summary = NULL)
plot(tab)

@@

```{r abundance}

# Abundance table

# Absolute abundances for the single most abundant taxa in each sample
tab <- dominance(pseq, index = "all")
kable(head(tab))
stargazer(tab, type = "text",summary = NULL)
plot(tab)

# Abunance histogram

# Visualize population densities for specific taxa
plot_density(pseq, "Dialister") + ggtitle("Absolute abundance")

# Same with log10 compositional abundances
x <- microbiome::transform(pseq, "compositional")
tax <- "Dialister"
plot_density(x, tax, log10 = TRUE) +
  ggtitle("Relative abundance") +
  xlab("Relative abundance (%)")

```
```{r dominance}

dominance(pseq, index = "all", rank = 1, relative = TRUE, aggregate = TRUE)

@@

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
```{r rarity and low abundance}

# Rarity and low abundance
tab <- rarity(pseq, index = "all")
kable(head(tab))
stargazer(tab, type = "text",summary = NULL)
plot(tab)

```
```{r core abundance}

# Core abundance
# relative proportion of the core species
tab <- core_abundance(pseq, detection = 0.1/100, prevalence = 50/100)
kable(head(tab))
stargazer(tab, type = "text",summary = NULL)
plot(tab)

```{r gini index}

# Gini index
# measure for relative inequality in economical income
# measures community diversity
# Gini index is between 0-1
# increasing gini index implies increasing inequality
tab <- inequality(pseq)
kable(head(tab))
stargazer(tab, type = "text",summary = NULL)
plot(tab)

```
```{r evenness}

# Evenness
tab <- evenness(pseq, "all", zeroes = TRUE)
kable(head(tab))
stargazer(tab, type = "text",summary = NULL)
plot(tab)

```
```{r prepare data for visualisation}

# Prepare data for vizualisation

# Get metadata
ps1.meta <- meta(ps1)
kable(head(ps1.meta))
stargazer(ps1.meta, type = "text",summary = NULL)

# Add the diversity table to metadata
ps1.meta$Shannon <- tab$shannon 
ps1.meta$InverseSimpson <- tab$inverse_simpson

# Compare differences in Shannon index between bmi group of the study subjects

# Create a list of pairwise comaprisons
bmi <- levels(ps1.meta$bmi_group) # get the variables

# Make a pairwise list that we want to compare.
bmi.pairs <- combn(seq_along(bmi), 2, simplify = FALSE, FUN = function(i)bmi[i])

print(bmi.pairs)

```{r violin plot}

# Violin plot
p1 <- ggviolin(ps1.meta, x = "bmi_group", y = "Shannon",
 add = "boxplot", fill = "bmi_group", palette = c("#a6cee3", "#b2df8a", "#fdbf6f","#a6cee3", "#b2df8a", "#fdbf6f"))

print(p1)

```{r add mean comparison p-values to a ggplot}

# Add mean comparison p-values to a ggplot
p1 <- p1 + stat_compare_means(comparisons = bmi.pairs) 
print(p1)

```
# https://rpkgs.datanovia.com/ggpubr/index.html
# Package for elegant data visualization in R

```{r coverage}

# Coverage

# The coverage index gives the number of groups needed to have a given proportion of the ecosystem occupied (by default 0.5 ie 50%).

tab <- coverage(pseq, threshold = 0.5)
kable(head(tab))
stargazer(tab, type = "text",summary = NULL)
plot(tab)

MICROBIOME COMPOSITION

```{r composition}

# Just use prevalent taxa to speed up examples
# (not absolute counts used in this example)

# Pick sample subset
library(phyloseq)

library(hrbrthemes)
library(gcookbook)
library(tidyverse)
library(dplyr)
library(microbiome)

######## THIS ALTERED THE CORE DATA ###############
#pseq <- core(atlas1006, detection = 10^3, prevalence = 95/100)
#pseq2 <- subset_samples(pseq, group == "DI" & nationality == "AFR" & timepoint.within.group == 1)

# Try another theme
# from https://github.com/hrbrmstr/hrbrthemes

# Limit the analysis on core taxa and specific sample group
p <- plot_composition(pseq,
              taxonomic.level = "OTU",
                      sample.sort = "nationality",
                      x.label = "nationality") +
     guides(fill = guide_legend(ncol = 1)) +
     scale_y_percent() +
     labs(x = "Samples", y = "Relative abundance (%)",
                                   title = "Relative abundance data",
                                   subtitle = "Subtitle",
                                   caption = "Caption text.") + 
     theme_ipsum(grid="Y")
print(p)

# Averaged by group
p <- plot_composition(pseq,
                      average_by = "bmi_group", transform = "compositional")
print(p)

```
```{r composition heatmap}

# Composition heatmap
tmp <- plot_composition(pseq, plot.type = "heatmap", transform = "compositional", 
            sample.sort = "neatmap", otu.sort = "neatmap", mar = c(6, 13, 1, 1))

print(tmp)

```{r plot taxa prevalence}

# Plot taxa prevalence
# overview of OTU prevalences alongwith their taxonomic affiliations
# will aid in checking if you filter OTUs based on prevalence, then what taxonomic affliations will be lost

data(atlas1006)

# Use sample and taxa subset to speed up example
p0 <- subset_samples(atlas1006, DNA_extraction_method == "r")

# Define detection and prevalence thresholds to filter out rare taxa
p0 <- core(p0, detection = 10, prevalence = 0)

# For the available taxonomic levels
plot_taxa_prevalence(p0, "Phylum", detection = 10)

```{r}

CORE MICROBIOME
```{r load example data for core microbiome}

# Load data
library(microbiome)
data(peerj32)

# Rename the data
pseq <- peerj32$phyloseq

# Calculate compositional version of the data
# (relative abundances)
pseq.rel <- microbiome::transform(pseq, "compositional")

```{r prevalence of taxonomic groups}

# Prevalence of taxonomi groups

# Relative population frequencies; at 1% compositional abundance threshold
head(prevalence(pseq.rel, detection = 1/100, sort = TRUE))

# Absolute population frequencies (sample count)
head(prevalence(pseq.rel, detection = 1/100, sort = TRUE, count = TRUE))

```{r core analysis}

# Core microbiota analysis

# Names of core taxa
core.taxa.standard <- core_members(pseq.rel, detection = 0, prevalence = 50/100)

# Full phyloseq object of core data
pseq.core <- core(pseq.rel, detection = 0, prevalence = .5)

# Retrieving the associated taxa names from the phyloseq object
core.taxa <- taxa(pseq.core)

# Total core abundance in each sample (sum of abundances of the core members)
core.abundance <- sample_sums(core(pseq.rel, detection = .01, prevalence = .95))

CORE VISUALIZATION
```{r core line plots}

# Core line plots

# Determine core microbiota across various abundance/prevalence thresholds with the blanket analysis

# With compositional (relative) abundances
det <- c(0, 0.1, 0.5, 2, 5, 20)/100
prevalences <- seq(.05, 1, .05)
plot_core(pseq.rel, prevalences = prevalences, detections = det, plot.type = "lineplot") + xlab("Relative Abundance (%)")

```{r core heat maps}

# Core heat maps

# Core with compositionals:
prevalences <- seq(.05, 1, .05)
detections <- 10^seq(log10(1e-3), log10(.2), length = 10)

# Also define gray color palette
gray <- gray(seq(0,1,length=5))
p <- plot_core(pseq.rel, plot.type = "heatmap", colours = gray,
    prevalences = prevalences, detections = detections) +
    xlab("Detection Threshold (Relative Abundance (%))")
print(p)

# Core with absolute counts and horizontal view:
# and minimum population prevalence (given as percentage)
detections <- 10^seq(log10(1), log10(max(abundances(pseq))/10), length = 10)

library(RColorBrewer)
p <- plot_core(pseq, plot.type = "heatmap", 
             prevalences = prevalences,
             detections = detections,
         colours = rev(brewer.pal(5, "Spectral")),
         min.prevalence = .2, horizontal = TRUE)
print(p)

As it can be seen, we see only OTu IDs and this may not be useful to interpret the data. We need to repreoccess this figure to include taxonomic information. We can do this as follows:

library(RColorBrewer)
library(knitr)
# Core with absolute counts and vertical view:
# and minimum population prevalence (given as percentage)
detections <- 10^seq(log10(1), log10(max(abundances(pseq.2))/10), length = 10)

healthycore <- plot_core(pseq.2, plot.type = "heatmap", 
             prevalences = prevalences,
             detections = detections,
         colours = rev(brewer.pal(5, "Spectral")),
         min.prevalence = .90, horizontal = F)
# get the data used for plotting 
df <- healthycore$data

# get the list of OTUs
list <- df$Taxa

# check the OTU ids
# print(list)

# get the taxonomy data
tax <- tax_table(pseq.2)
tax <- as.data.frame(tax)

# add the OTus to last column
tax$OTU <- rownames(tax)

# select taxonomy of only 
# those OTUs that are used in the plot
tax2 <- dplyr::filter(tax, rownames(tax) %in% list)

# head(tax2)

# We will merege all the column into one except the Doamin as all is bacteria in this case
tax.unit <- tidyr::unite(tax2, Taxa_level,c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species", "OTU"), sep = "_;", remove = TRUE)

tax.unit$Taxa_level <- gsub(pattern="[a-z]__",replacement="", tax.unit$Taxa_level)

# add this new information into the plot data df

df$Taxa <- tax.unit$Taxa_level

# you can see now we have the taxonomic information
knitr::kable(head(df))

# replace the data in the plot object
healthycore$data <- df

plot(healthycore + theme(axis.text.y = element_text(face="italic")))

```{r}

MICROBIOME STABILITY ANALYSIS

Get example data - HITChip Atlas of 130 genus-like taxa across 1006 healthy western adults. A subset of 76 subjects have also short time series available for temporal stability analysis:
```{r microbiome stability analysis}

# Load the example data
set.seed(134)
library(microbiome)
library(dplyr)
data(atlas1006)

# Rename the example data
pseq <- atlas1006

# Focus on specific subset
pseq <- pseq %>% subset_samples(DNA_extraction_method == "r")

# Use relative abundances
pseq <- microbiome::transform(pseq, "compositional")

# Keep only the prevalent taxa to speed up examples
pseq <- core(pseq, detection = .1/100, prevalence = 99/100)

# For cross-sectional analysis, use only the baseline time point:
pseq0 <- baseline(pseq)

It has been reported that certain microbial groups exhibit bi-stable abundance distributions with distinct peaks at low and high abundances, and an instable intermediate abundance range. Instability at the intermediate abundance range is hence one indicator of bi-stability. Lahti et al. 2014 used straightforward correlation analysis to quantify how the distance from the intermediate abundance region (50% quantile) is associated with the observed shifts between consecutive time points.
```{r intermediate stability quantification}

# Intermediate stability quantification
intermediate.stability <- intermediate_stability(pseq, output = "scores")

BIMODALITY ANALYSIS

Bimodality of the abundance distribution provides an indirect indicator of bistability, although other explanations such as sampling biases etc. should be controlled. Multiple bimodality scores are available.
```{r}

# Load the example data
library(microbiome)
library(dplyr)
data(atlas1006)

# Rename the example data
pseq <- atlas1006

# Focus on specific DNA extraction method
pseq <- pseq %>% subset_samples(DNA_extraction_method == "r")

# Keep prevalent taxa (HITChip signal >3 in >20 percent of the samples)
pseq <- core(pseq, detection = 10^3, prevalence = .2)

# Use relative abundances
pseq <- microbiome::transform(pseq, "compositional")

# For cross-sectional analysis, include
# only the zero time point:
pseq0 <- subset_samples(pseq, time == 0)

# Bimodality is better estimated from log10 abundances
pseq0.log10 <- microbiome::transform(pseq0, "log10")
bimodality <- bimodality(pseq0.log10, method = "potential_analysis", bs.iter = 20)

```{r bimodality visualization}

# Visualize population densities for unimodal and bimodal groups

# Pick the most and least bimodal taxa as examples
unimodal  <- names(sort(bimodality))[[1]]
bimodal  <- rev(names(sort(bimodality)))[[1]]

# Visualize population frequencies
library(ggplot2)
theme_set(theme_bw(20))
p1 <- plot_density(pseq, variable = unimodal, log10 = TRUE) 
p2 <- plot_density(pseq, variable = bimodal,  log10 = TRUE) 
library(gridExtra)
library(ggplot2)
grid.arrange(p1, p2, nrow = 1)

```{r tipping point detection}

# Tipping point detection

# Log10 abundance for a selected taxonomic group
tax <- bimodal

# Detect tipping points detection at log10 abundances 
x <- log10(abundances(pseq)[tax,])

# Bootstrapped potential analysis to identify potential minima
set.seed(3432)
potential.minima <- potential_analysis(log10(abundances(pseq)[tax,]), bs.iter = 50)$minima
# Same with earlywarnings package (without bootstrap ie. less robust)
# library(earlywarnings)
# res <- livpotential_ews(x)$min.points

# Identify the potential minimum location as a tipping point candidate
# and cast the tipping back to the original (non-log) space:
tipping.point <- 10^potential.minima

# Identify potential minima in cross-section population data as tipping point candidates.

# Log10 abundance for a selected taxonomic group
# Pick the most bimodal taxa as an example
tax  <- names(which.max(bimodality.score))

# Detect tipping points detection at log10 abundances 
x <- abundances(microbiome::transform(pseq, "log10"))[tax,]

# Bootstrapped potential analysis to identify potential minima
# in practice, use more bootstrap iterations
potential.minima <- potential_analysis(x, bs.iter = 10)$minima

# Same with earlywarnings package (without bootstrap ie. less robust)
# library(earlywarnings)
# res <- livpotential_ews(x)$min.points

# Identify the potential minimum location as a tipping point candidate
# and cast the tipping back to the original (non-log) space:
tipping.point <- 10^potential.minima

print(tipping.point)

Pick subset of the HITChip Atlas data set and plot the subject abundance variation lineplot (Variation tip plot) and Bimodality hotplot for a given taxon as in Lahti et al. 2014. The Dialister has bimodal population distribution and reduced temporal stability within subjects at intermediate abundances.
```{r variation lineplot and bimodality hotplot}

# Variation lineplot and bimodality hotplot

# We set the tipping point manually in this example

# Bimodality hotplot:
# Consider a unique sample from each subject: the baseline time point 
p <- hotplot(pseq0, tax, tipping.point = 0.005)
print(p)

# Visualize bimodality
pv <- plot_tipping(pseq, tax, tipping.point = 0.005)
print(pv)

Comparing bimodality and intermediate stability
The analysis suggests that bimodal population distribution across individuals is often associated with instable intermediate abundances within individuals. The specific bi-stable groups in the upper left corner were suggested to constitute bistable tipping elements of the human intestinal microbiota in Lahti et al. Nat. Comm. 5:4344, 2014:
```{r comparing bimodality and intermediate stability}

# Comparing bimodality and intermediate stability

taxa <- taxa(pseq0)
df <- data.frame(group = taxa,
                 intermediate.stability = intermediate.stability[taxa],
         bimodality = bimodality.score[taxa])

theme_set(theme_bw(20))
p <- ggplot(df,
       aes(x = intermediate.stability, y = bimodality, label = group)) +
       geom_text() +
       geom_point() 
print(p)

```{r}
